From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex
Fixes and fixes

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{Carette2016} and~\citet{threemodels} and established in this work.

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case).

% We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
% $\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
% doesn't have one, but we could show externally that it satisfies univalence.

Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

%%%%%%%%%%%%%%
% \paragraph{Univalent Fibrations}

% Coherence problems in type theory, coherence via Well-Foundedness.

% Formalised proofs of Mac Lane's coherence theorem.

% Applications of FSMG and history of the coherence theorems.

% Other proofs of coherence theorems, Joyal-Street.

% Pi has other extensions (fractional/negative/recursive types).
% What are the free X monoidal structures they're describing?

% Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
% should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.

% Other applications of symmetric groups.

% Actions of symmetric groups, permutation groupoids.

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages}

@@

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
established in this work.

@@

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
constructions are hard (or impossible in the general case).

We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
doesn't have one, but we could show externally that it satisfies univalence.

%%%%%%%%%%%%%%

Coherence problems in type theory, coherence via Well-Foundedness.

Formalised proofs of Mac Lane's coherence theorem.

Applications of FSMG and history of the coherence theorems.

Other proofs of coherence theorems, Joyal-Street.

Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.

Other applications of symmetric groups.

Actions of symmetric groups, permutation groupoids.

### 2021-07-09 Pi+/paper/finite.tex
Fix crefs
Fixes and fixes

We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
(in~\Cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
first.

@@

  \begin{definition}~\label{def:group}
    In type theory, a group $G$ can be defined as a set $S$ with the following pieces of data:

@@

\begin{propositionrep}~\label{prop:fin-finexcept}
  \leavevmode
  \begin{enumerate}
    % \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-1}
    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
    \item For any $n : \Nat$, \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
            \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
  \end{enumerate}
\end{propositionrep}

@@

\begin{proposition}~\label{prop:lehmer-aut-equiv}
  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
\end{proposition}

>>>>

We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
first.

@@

  \begin{definition}
  \label{def:group}
    In type theory, a group $G$ can be defined as a set $S$ with the following pieces of data:

@@

\begin{propositionrep}
  \label{prop:fin-finexcept}
  \leavevmode
  \begin{enumerate}
    % \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-1}
    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
    \item For any $n : \Nat$, \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
            \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
  \end{enumerate}
\end{propositionrep}

@@

\begin{proposition}
  \label{prop:lehmer-aut-equiv}
  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
\end{proposition}

### 2021-07-09 Pi+/paper/ufin.tex
Fixes and fixes

% Fourth - explain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. (See App.~\ref{app:grpdexample} for an example on a 3-element set.) This is the
groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
and proceed to construct the particular instance for finite types, $\UFin$ (\cref{def:ufin}).

@@

%%%
\subsection{The Type Theory}~\label{subsec:type-theory}
% \subsection{Univalent Subuniverses}
% \label{sec:univalent}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well.  In App.~\ref{app:identitytypes}, App.~\ref{app:homotopytypes}, and App~\ref{app:hits}, we review the basics of
identity types, homotopy types, and HITs and refer the reader to the book for more details.

@@

  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
  the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
  universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called the
  fiber over $x$.  The $\term{transport}$ operation, named $\term{tr}$, lifts paths in the indexing type to functions
  between fibers.
  %which is an $A$-indexed family of groupoids.

  \begin{gather*}
    \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
    \qquad
    \term{tr}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
  \end{gather*}

@@

  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}

@@

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration. The
$\term{transport}$ operation $\transport{P}{p}{\blank}$ lifts paths in the base space to equivalences between the fibers:
% Using the groupoid structure of
% $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence.
% We name
% this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

@@

The use of the word \emph{univalent} here is a reference to Voevodsky's \emph{univalence} principle. Indeed, univalence
characterises paths in the universe as equivalences between types, which follows from the canonical fibration
$\idfunc : \UU \to \UU$ being univalent.

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent and generates a
  univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.
\end{propositionrep}

@@

The types we are interested in are the finite types. In constructive mathematics, the notion of finiteness is
subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is merely
equivalent to a finite set (\cref{def:finite-set} and \cref{def:isfin}).

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.

@@

\subsection{Rig structure}~\label{subsec:rig}

@@

\begin{proposition}
  For any $n, m : \Nat$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot        \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top        \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]
    \end{aligned}
  \end{gather*}
  and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \bot \sqcup X         & \eqv X                     \\
      X \sqcup \bot         & \eqv X                     \\
      (X \sqcup Y) \sqcup Z & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y            & \eqv Y \sqcup Y            \\
      X \times \bot         & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \top \times X         & \eqv X                                \\
      X \times \top         & \eqv X                                \\
      (X \times Y) \times Z & \eqv X \times (Y \times Z)            \\
      X \times Y            & \eqv Y \times X                       \\
      X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

>>>>

% Fourth - explain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the
groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
and proceed to construct the particular instance for finite types, $\UFin$ (\cref{def:ufin}).

@@

%%%
\subsection{The Type Theory}
\label{subsec:type-theory}
% \subsection{Univalent Subuniverses}
% \label{sec:univalent}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
as well.  In~\cref{app:identitytypes},~\cref{app:homotopytypes}, and~\cref{app:hits}, we review the basics of identity
types, homotopy types, and HITs and refer the reader to the book for more details.

@@

  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action
  on the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to
  the universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called
  the fiber over $x$.  The $\term{transport}$ operation, named $\term{transport}/term{tr}$, lifts paths in the indexing
  type to functions between fibers.
  %which is an $A$-indexed family of groupoids.

  \begin{gather*}
    \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
    \qquad
    \term{transport}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
  \end{gather*}

@@

  \begin{figure}
    \begin{center}
      \begin{tikzpicture}[yscale=.5,xscale=2]
        \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
        \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
        \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
        \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
        \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
        \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
        \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
      \end{tikzpicture}
    \end{center}
    \caption{Lifting operation in $P$}
    \label{fig:lift}
  \end{figure}

@@

Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
the base space to functions between the fibers.  Using the groupoid structure of $A$, for a path $p : x \id_{A} y$,
$\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
\[
  \tptEqv{P} : \dfun*{x,y:A}{(x \id_{A} y) \to (P(x) \eqv P(y))}
\]

@@

Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018}, to build a model of Voevodsky's
\emph{univalence} principle in simplicial sets. Indeed, univalence characterises paths in the universe as equivalences
between types, which follows from the canonical fibration $\idfunc : \UU \to \UU$ being univalent.

@@

\begin{propositionrep}[Univalent Subuniverse]
  \label{prop:univsub}
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent
  and generates a univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.~\footnote{Univalent
    typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent subuniverses.}
\end{propositionrep}

@@

The types we are interested in are the finite types. In constructive mathematics, the notion of finiteness is
subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is merely
equivalent to a finite set (\cref*{def:finite-set,def:isfin}).

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
\footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
  by~\citet{christensenCharacterizationUnivalentFibrations2015}.}

@@

\subsection{Rig structure}
\label{subsec:rig}

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,

  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
      \\
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup Y            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
      \\
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}
